Detecting Network Motifs in Knowledge Graphs using Compression

### 2017-08-10 kgmotifs/kgmotifs.tex
...

\begin{abstract}
We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We introduce a scalable approach for detecting such motifs in large knowledge graphs, inspired by recent work for simple graphs, and show that the motifs returned reflect the basic structure of the graph. Specifically, we show that common motifs reflect graph patterns used in common queries, basic schematic units of the graph and meaningful functional subunits, for various knowledge graphs.
\end{abstract}

\noindent The Linked, Open Data cloud contains a wealth of knowledge graphs, and is growing quickly. At the time of writing the average knowledge graph contains over 65 000 edges, with many datasets containing more than a million.\footnotemark~For such large graphs, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of objects? What are small, recurring patterns that might represent a novel insight into the data?

@@

In \cite{bloem2017large}, an alternative method is presented that uses compression as a heuristic for motif value: the better a motif compresses the data, the more likely it is to be meaningful. It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to classical motif analysis. Crucially, this method scales to graphs with billions of edges, suggesting that it might be suitable for the scale of knowledge graphs we encounter in the LOD cloud.

@@

@@

A \emph{knowledge graph} $G$, hereafter known simply as a \emph{graph} is a, is tuple $G = (V_G, E_G)$ with finite sets $V_G \subset V$ and $E_G \subset V \times R \times V$. We define the triple list of $G$ as $G^T = \{(s, p, o) \in \N^3 : \left (v(s), r(p), v(o) \right ) \in E_G\}$. In this paper we will \emph{identify} the graph with the triple list. That is when we speak of encoding the graph, we mean encoding the triple list, and when we speak of a probability distribution over graphs, we mean a probability distribution over triple lists. Let $R_G \cub \N$ be the set of relations used in the $G$.

This means that we are purposely \emph{not} modelling certain aspects of the data. There might, for instance be a similarity between two node labels $v(a)$ and $v(b)$ (such as a shared IRI prefix). We are not interested in exploiting such structure, so we eliminate it from our notation. Additionally, we introduce some bias with our choice of $e$ and $r$. If we were to use a simple code to store the integers in the triple-list we will end up using fewer bits for those nodes and triples with labels who were assigned short integers. Equivalently, when sampling from a simple distribution over the integers, small numbers will be more likely to be generated. In practice, however, we will always \emph{learn} a distribution over the integers from the data, so that $e$ and $r$ are purely a means to simplify the notation. Which functions we choose will not affect our model.

As shown in Figure~\ref{figure:diagram}, the motifs that we aim to find are partially labeled: some nodes and edges are labeled with the marker "?" indicating that when we fit the motif to the graph, those edges can contain any label. In other words, the motif is a \emph{basic graph pattern}, and the edges and nodes marked with "?" are variables. To represent this, we add a single variable symbol $?$. Thus, a motif $M$ is defined by the triple set $M \subset (? \cup \N) ^3$, with integer values in triples interpreted as they are in complete knowledge graphs.

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of \emph{instances} $\cal I'$ where $I \in \cal I'$ is a subset of $E_G$ which matches the pattern. Moverover, we require that all instances in $\cal I$ are mutually disjoint. Thus, we cannot simply return all matches of the pattern, we require some selection to ensure that none of them overlap. In most settings we do not have a given $M$ and $I$ and we will need to search for them. We describe a simple and fast search algorithm in Section~\ref{section:search}.

@@

% TODO

@@

\begin{description}
 \item[the graph dimensions] We first store $|V_G|$, $|E_G|$ and $|R_G|$ using the generic code $L^{\N}(\cdot)$.
 \item[the pattern] We store the pattern $M$ using $L^\text{base}$. We increment each of the pattern's existing link labels by 1, and use the label 0 to represent unlabeled links. Since $L^\text{base}$ does not store node labels ($V_M$ is always the set of the first $|V_M|$ natural numbers) we need to store separately to which nodes in $G$ the labeled nodes of $M$ refer. We store a sequence of integers ${g_i}$ where $g_i = 0$ indicates that the $i$-th node of $M$ is unlabeled, and $g_i \neq 0$ indicates that $g_i$ should map to node $g_i - 1$ in $G$.
 \item[the template] This is the graph, minus all links occurring in  instances. Let $E_G'$ be $E_G$ minus any link occurring in any member of $\cal I$. We then store $(V_G, E_G')$ using $L^\text{base}$.
 \item[the instances] In order to reconstruct the graph from the template, we need to know which nodes correspond to instances of the motif, and in what order. For each unlabeled node in M, we store one list of length $\cal I$ of the nodes in $G$ to which it maps for each instance. We store these lists using the DM code, since it is likely that a particular subset of the nodes of $G$ is frequent for a particular unlabeled node of a pattern.
 \item[the edge labels] We now have enough information to reconstruct the graph structure. All that remains to fully reconstruct $G$ is to label the unlabeled link of the motif instances. As for the nodes, we create one list of length $\cal I$ for each unlabeled  link in the pattern which we store using the DM code.
\end{description}

We can note a few aspects of this code. Firstly, it achieves its compression from not having to store the edges of the motif instances. For each additional instance, it pays the cost of storing the instance, and any unlabeled nodes and links, but it saves $E_M$ edges. The precise trade-off depends on the choice of base code, and on the structure of the rest of the graph (some edges are cheaper to store than others).

Nevertheless, we can make a few broad observations: patterns with many unlabeled nodes and links are more costly to store than fully labeled patterns. On the other hand, a pattern with many unlabeled nodes and links is more likely to have many instances, so that \emph{if} it results in a small amount compression per instance, the large number of instances may cause it to compress a lot. For the unlabeled nodes and links, we use a DM model for each specific node and link. This means that the more structure there is to the label given to a particular part of the pattern, the better we compress. For instance, suppose a particular node of the pattern is always labeled with either a node corresponding to YES or NO, in equal proportion, we will require only a bit per instance to record those labels. If all nodes in the graph are equally likely to appear in that position, we will pay approximately $\log |V_G|$ bits per instance. If only a single node appears in the position, we pay only a a constant number of bits to record the entire sequence of instances. \footnotemark

\footnote{In fact, we could simplify the model by keeping the pattern entirely unlabeled, and trusting that the DM code will store everything efficiently. However, for our search algorithm, and for the interpretation of the motif as a graph pattern, we prefer to distinguish between labeled and unlabeled elements explicitly.}

@@

In most use cases, we do not start with a given motif and a set of instances. We have only the graph $G$, and we wish to find any motifs that might be interesting. In order to do this efficiently, we generate candidate patterns by \emph{sampling} them from the data. This method, inspired by \cite{kashtan2004efficient} is based on the simple insight that when we sample a pattern from the data, we are more likely to sample a frequently occurring pattern than a infrequent one. In we sample $K$ times, and rank the results by relative frequency within the sample, we are likely to end up with a good indication of the most frequent patterns in the data. For unlabeled graphs, as little as 500 samples can be sufficient to find the most frequent subgraph.

Theoretically, this allows us to generate promising motifs with time complexity independent of the size of the graph. Our algorithm samples a pattern in three steps. First, it samples a weakly connected subset of $n$ nodes, and extracts the subgraph induced by these nodes. Second, it converts this subgraph to a non-induced pattern by randomly removing labels from some of the nodes and links and randomly removing some links entirely (as detailed below). Finally, we sort the nodes of this pattern into a canonical ordering, using the Nauty algorithm \cite{mckay1981practical}.

For a subgraph with $n$ nodes and $e$ links, we can generate $n^2e^3$ patterns. A uniform random choice among these would make patterns with many labels very unlikely. Instead, we sample $n' ~ U([0.. n])$, $e' ~ U([0.. e])$ and $e^*~ U([0..e'])$: the number of nodes to leave labeled, the number of nodes links to leave, and the number of links to leave labeled respectively. If the resulting pattern is disconnected, we reject the sample.

@@

\footnotemark{In \cite{bloem2017large}, the number of edges between nodes inside the instance and outside it (the exdegree of the instance), was a strong predictor of whether the instances would benefit compression. Thus, the algorithm was strongly helped by pruning the list of instances using a ternary search. For the current code, the exdegree does not affect compression (since the instance nodes are not removed from the template graph). Therefore, we pass the entire instance list $\I_M$ to the relevance test wihtout pruning it.}

@@

\begin{figure}[bt]
{
\hrule
\vspace{3mm}
\raggedright
Given: a graph $G$, \\
\tab a minimum motif size $s_\text{min}$ and a maximum motif size $s_\text{max}$ \\
\tab a target number of graphs sampled\\
\tab\\
Initialize $d$, a dictionary from motif $M$ to a list of instances $d[M] \in 2^{2^N}$ \\ 
  \\
Choose $s$ uniformly from $[s_\text{min} .. s_\text{max}]$ \\
Choose a starting node $n$ uniformly from $G$ \\
$N \from \{n\}$ \\
\textbf{while} $|N| < s$:\\
\tab Choose $n$ uniformly from $N$
\tab Choose $n'$ uniformly from the neighbors of $n$
\tab $N \from N \cup n'$ \\
\tab After 500 cycles, sample a different starting node\\

>>>>

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We introduce a scalable approach for detecting such motifs in large knowledge graphs, inspired by recent work for simple graphs, and show that the motifs returned reflect the basic structure of the graph. Specifically, we show that common motifs reflect graph patterns used in common queries, basic schematic units of the graph and meaningful functional subunits, for various knowledge graphs.
\end{abstract}

\noindent The Linked Open Data cloud contains a wealth of knowledge graphs, and is growing quickly. At the time of writing the average knowledge graph contains over 65\,000 edges, with many datasets containing more than a million.\footnotemark~For such large graphs, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of objects? What are small, recurring patterns that might represent a novel insight into the data?

@@

In \cite{bloem2017large}, an alternative method is presented that uses \emph{compression} as a heuristic for motif value: the better a motif compresses the data, the more likely it is to be meaningful. It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to classical motif analysis. Crucially, this method scales to graphs with billions of edges, suggesting that it might be suitable for the scale of knowledge graphs we encounter in the LOD cloud.

@@

% TODO: General purpose codes: natural numbers, integers, sequences.

@@

A \emph{knowledge graph} $G$, hereafter known simply as a \emph{graph} is a, is tuple $G = (V_G, E_G)$ with finite sets $V_G \subset V$ and $E_G \subset V \times R \times V$. We define the \emph{tripleset} of $G$ as $G^T = \{(s, p, o) \in \N^3 : \left (v(s), r(p), v(o) \right ) \in E_G\}$. In this paper we will \emph{identify} the graph with the tripleset. That is, when we speak of encoding the graph, we mean encoding the triple list, and when we speak of a probability distribution over graphs, we mean a probability distribution over triplesets. Let $R_G \subset \N$ be the set of relations used in $G$.

This means that we are purposely \emph{not} modelling certain aspects of the data. There might, for instance be a similarity between two node labels $a$ and $b$ (such as a shared IRI prefix). We are not interested in exploiting such structure, so we eliminate it from our notation. We also introduce some bias with our choice of $e$ and $r$. Entities and relations that are assigned small integers may end up receiving short codes or high probability mass, if a simple code or disgribution is used. In practice, however, we will always \emph{learn} a distribution over the integers from the data, so that $e$ and $r$ are purely a means to simplify the notation. Which functions we choose will not affect our model.

As shown in Figure~\ref{figure:diagram}, the motifs that we aim to find are partially labeled: some nodes and edges are labeled with the marker "?" indicating that when we fit the motif to the graph, those edges can contain any label. In other words, the motif is a \emph{basic graph pattern}, and the edges and nodes marked with "?" are variables. To represent this, we require additional symbols: one for variable relations, and one for each node in the motif marked with a variable. We will define a pattern by an \emph{integer} triple list: $M \subset \Z^3$. Natural numbers mean the same as they do in a graph, and negative integers represent variable nodes. Variable links marked with the same negative integer, must map to the same relation in order for the pattern to match, but variable links labeled with different negative integers \emph{may} map to the same relation. For the current paper, we will only search for patterns where each variable link contains a unique label.

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of \emph{instances} $\cal I'$ where $I \in \cal I'$ is a subset of $E_G$ which matches the pattern. Moreover, we require that all instances in $\cal I$ are mutually disjoint. Thus, we cannot simply return all matches of the pattern, we require some selection to ensure that none of them overlap. In most settings we do not have a given $M$ and $I$ and we will need to search for them. We describe a simple and fast search algorithm in Section~\ref{section:search}.

Note that unlike the graph, the nodes of $M$ \emph{are} explicitly labeled. The positive-valued nodes of $M$ are not simply the first $|V_M|$ integers, but refer specifically to the nodes of $G$.

@@

@@

\begin{description}
 \item[the graph dimensions] We first store $|V_G|$, $|E_G|$ and $|R_G|$ using the generic code $L^{\N}(\cdot)$.
 \item[the pattern] Let $M'$ be the unlabeled graph representation of $M$: that is, $M'$ is a plain graph (not a pattern), $V_{M'}$ contains only the first $V_{M'}$ natural numbers and similarly for $R_{M'}$. Otherwise, $M$ is isomorphic to $M'$: the node and edge labels of $M$ can be mapped to those of $M'$ so that $M$ only contains an edge $E$ if and only if $M'$ contains the edge resulting from mapping $E$'s symbols. 
 \item[the template] This is the graph, minus all links occurring in  instances. Let $E_G'$ be $E_G$ minus any link occurring in any member of $\cal I$. We then store $(V_G, E_G')$ using $L^\text{base}$.
 \item[the instances] In order to reconstruct the graph from the template, we need to know which nodes correspond to instances of the motif, and in what order. For each node in $M'$, we create a list of length $\cal I$ of the nodes in $G$ to which it maps for each instance. For the non-variable list, this will be a a list simply repeating one node, and for the variable nodes it will be a list containing some subset of the ndoes in $G$.
 We store each list using the DM code. It may seem wasteful to store a list repeating a single element, but under the DM code such lists will be stored efficiently. The more varied the list, the more bits it will require to store.
 \item[the edge labels] We now have enough information to reconstruct the graph structure. All that remains to fully reconstruct $G$ is to label the links of the motif instances. As for the nodes, we create one list of length $\cal I$ for each type of link in $M'$, which we store using the DM code.
\end{description}

The precise computation of the codelength is given in Algorithm~\ref{algorithm:codelength}.

We note a few aspects of this code. Firstly, it achieves its compression by not storing the edges of the motif instances. For each additional instance, it pays the cost of storing the instance, and any variable nodes and links, but it saves $|E_M|$ edges in storing the template. The precise trade-off depends on the choice of base code, and on the structure of the rest of the graph (some edges are cheaper to store than others).

Nevertheless, we can make a few broad observations: patterns with many variable nodes and links are more costly to store than fully labeled patterns. On the other hand, a pattern with many variable nodes and links is more likely to have many instances, so that \emph{if} it results in a small amount compression per instance, the large number of instances may cause it to compress a lot. For the variable nodes and links, we use a DM model for each specific node and link. This means that the more structure there is to the label given to a particular part of the pattern, the better we compress. For instance, suppose a particular node of the pattern is always labeled with either a node corresponding to YES or NO, in equal proportion, we will require only a bit per instance to record those labels. If, however, all nodes in the graph are equally likely to appear in the position of the variable, we will pay approximately $\log |V_G|$ bits per instance. If only a single node appears in the position, we pay only a a constant number of bits to record the entire sequence of instances. Thus, to our code, non-variable nodes are just a highly regular type of variable node. The distinction is only meaningful in the search algorithm.

@@

In most use cases, we do not start with a given motif and a set of instances. We have only the graph $G$, and we wish to find any motifs that might be interesting. In order to do this efficiently, we generate candidate patterns by \emph{sampling} them from the data. This method, inspired by \cite{kashtan2004efficient} is based on the simple insight that when we sample a pattern from the data, we are more likely to sample a frequently occurring pattern than a infrequent one. Ik we sample $K$ times, and rank the results by relative frequency within the sample, we are likely to end up with a good indication of the most frequent patterns in the data. For unlabeled graphs, as few as 500 samples can be sufficient to find the most frequent subgraph.

Theoretically, this allows us to generate promising motifs with time complexity independent of the size of the graph. Our algorithm samples a pattern in three steps. First, it samples a weakly connected subset of $n$ nodes, and extracts the subgraph induced by these nodes. Second, it converts this subgraph to a non-induced pattern by randomly removing labels from some of the nodes and links and randomly removing some links entirely (as detailed below). Finally, we sort the nodes of this pattern into a canonical ordering, using the Nauty algorithm \cite{mckay1981practical}, this allows us to record the instances for each subgraph without counting isomorphic subgraphs separately.

For an induced subgraph with $n$ nodes and $e$ links, we can generate $n^2e^3$ patterns. A uniform random choice among these would make patterns with many labels very unlikely. Instead, we sample $n' ~ U([0.. n])$, $e' ~ U([0.. e])$ and $e^*~ U([0..e'])$, with $n'$ the number of nodes to leave labeled, $e'$ the number of links to leave, and $e^*$ the number of links to leave labeled, respectively. If the resulting pattern is disconnected, we reject the sample.

@@

\footnotemark{In \cite{bloem2017large}, the number of edges between nodes inside the instance and outside it (the exdegree of the instance), was a strong predictor of whether the instances would benefit compression. Thus, the algorithm was strongly helped by pruning the list of instances using a ternary search. For the current code, the exdegree does not affect compression (since the instance nodes are not removed from the template graph). Therefore, we pass the entire instance list $\I_M$ to the relevance test without pruning it.}

@@

\begin{figure}[bt]
{
\hrule
\vspace{3mm}
\raggedright
Given: a graph $G$, \\
\tab a minimum motif size $s_\text{min}$ and a maximum motif size $s_\text{max}$ \\
\tab a target number of graphs sampled\\
\tab\\
Initialize $d$, a dictionary from motif $M$ to a list of instances $d[M] \in 2^{2^N}$ \\ 
\tab\\
Choose $s$ uniformly from $[s_\text{min} .. s_\text{max}]$ \\
Choose a starting node $n$ uniformly from $G$ \\
$N \from \{n\}$ \\
\textbf{while} $|N| < s$:\\
\tab Choose $n$ uniformly from $N$
\tab Choose $n'$ uniformly from the neighbors of $n$
\tab $N \from N \cup n'$ \\
\tab After 500 cycles, sample a different starting node\\
